Modal Verification Patterns for Systems

### 2025-06-16 paper.tex
+

\usepackage{microtype}%if unwanted, comment out or use option "draft"
%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
%\usepackage{t1enc}
%\usepackage[strings]{underscore}
%%%%%\usepackage{geometry}

\usepackage[rawfloats=true]{floatrow} 
\restylefloat{figure}    
\DeclareMathSizes{4}{4}{4}{4}
\usepackage{varwidth}
\usepackage{lmodern}
\usepackage{amsfonts}
\usepackage[fleqn]{amsmath}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{proof}
\usepackage{stmaryrd}
\usepackage{xspace}
\usepackage{listings}
\usepackage{comment}
\usepackage{longtable}
\usepackage{float}
\usepackage[section]{placeins}
\usepackage[toc,page]{appendix}
\usepackage[export]{adjustbox}
\usepackage{amssymb}
\usepackage{graphics,graphicx}
\usepackage{mathpartir}
\usepackage{array}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric}
\usetikzlibrary{arrows.meta,arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{calc, shapes, backgrounds}
\usepackage{verbatim}
\usepackage{tikz-qtree}
\usepackage{pgfplots}
\usepackage{caption,capt-of}
\usepackage{color}
\usepackage{soul}
\setul{1ex}{0.8ex}
\definecolor{orange}{rgb}{1,0.5,0}
\setulcolor{orange}
\usepackage{subcaption}
\usepackage{lipsum}
\usepackage{afterpage}
%%sequeezeeee
\addtolength{\textfloatsep}{-10pt}
\addtolength{\dbltextfloatsep}{-10pt}

@@

\paragraph{Virtualization of Disk-Blocks in Filesystems} Likewise, 
\paragraph{Consistency of Disk Layout} \todo{}
\paragraph{Memory Reclamation in Client Code}\todo{iso:speak about Amal's stuff}
\paragraph{Consistency of Reclaimed Memory}\todo{iso: speak about the operator for which the hybrid logic stuff is usefull }

@@

\subsection{Virtualization}
One of the most common forms of abstraction provided by systems software is \emph{virtualization},
which abstracts the relationship between conceptual and physical computing resources. 
Operating System (OS) kernels virtualize memory locations and quantity (via virtual memory and paging~\cite{Denning1970VM}).
Distributed language runtimes may virtualize addresses, even when processes may migrate across machines~\cite{jul1988fine}.
Filesystems virtualize locations on disk~\cite{rodeh2013btrfs,hitz1994file,Rosenblum1992LFS,bonwick2003zettabyte}.
Programs built on top of the corresponding systems software layer work logically at the level of these virtualized resources,
and it makes sense to specify the systems software directly in terms of those abstractions.
\paragraph{Access by Translation}
Accessing virtualized resources via translation is a common way to virtualize notions of location (e.g.,
virtual memory addresses, inodes or object IDs instead of disk addresses~\cite{bonwick2003zettabyte,hitz1994file,Rosenblum1992LFS,rodeh2013btrfs}).
%Accessing the hardware resources in the presence of sharing means having more resource references (references to the virtualized resources) to access the \emph{mapped} resources that are physical and potentially fewer. 
B-trees, page tables, and related structures both behave like maps, when the corresponding physical resources exist as such
just in a different location.
Control over the lookup process (e.g, handling the case of a missing translation entry) allows for additional flexibility,
such as filling holes in sparse files, or demand paging (both from disk or lazily populating anonymous initially-zeroed mappings).
Although the realization of these maps may differ from a system to system based on the context (and sometimes hardware details),
they are semantically --- logically --- partial maps, worth treating as such in verification.

%\subsection{Sharing} Sharing is one of the fundamental reasons for kernels to {virtualize} resources,
%for creating the illusion of exclusive access to virtualized versions of shared resources, including CPU time (via preemption).
%An implementation of sharing relies on two concepts: \emph{indirection} and \emph{coordination}. Indirection is a method that allows the mapping of resources of one type into another. By doing so, it allows treating the limited resources (e.g. hardware resources such as physical memory) as if there are more than they actually are (e.g. via virtual memory addresses and paging). Consequently, virtualization for the sake of ensuring availability requires \emph{sharing} of the resources that exist in reality.
%%\paragraph{Isolation Hand-in-Hand with Sharing}
%%\Ismail{put a few words on abstractions such as address space, stack region, snapshots ,etc.}
%%\Colin{Not sure what you want here?}

>>>>

\usepackage{microtype}%if unwanted, comment out or use option "draft"
%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
%\usepackage{t1enc}
%\usepackage[strings]{underscore}
%%%%%\usepackage{geometry}
\usepackage{wrapfig}
\usepackage[rawfloats=true]{floatrow} 
\restylefloat{figure}    
\DeclareMathSizes{4}{4}{4}{4}
\usepackage{varwidth}
\usepackage{lmodern}
\usepackage{amsfonts}
\usepackage[fleqn]{amsmath}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{proof}
\usepackage{stmaryrd}
\usepackage{xspace}
\usepackage{listings}
\usepackage{comment}
\usepackage{longtable}
\usepackage{float}
\usepackage[section]{placeins}
\usepackage[toc,page]{appendix}
\usepackage[export]{adjustbox}
\usepackage{amssymb}
\usepackage{graphics,graphicx}
\usepackage{mathpartir}
\usepackage{array}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric}
\usetikzlibrary{arrows.meta,arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{calc, shapes, backgrounds}
\usepackage{verbatim}
\usepackage{tikz-qtree}
\usepackage{pgfplots}
\usepackage{caption,capt-of}
\usepackage{color}
\usepackage{soul}
\setul{1ex}{0.8ex}
\definecolor{orange}{rgb}{1,0.5,0}
\setulcolor{orange}
\usepackage{subcaption}
\usepackage{lipsum}
\usepackage{afterpage}
\usepackage{fontawesome}
\usepackage[utf8]{inputenc}

%%sequeezeeee
\addtolength{\textfloatsep}{-10pt}
\addtolength{\dbltextfloatsep}{-10pt}

\newcommand{\ptree}{\text{\faSitemap}@[\ell]}
\newcommand{\mtree}{\varupphi\text{\faSitemap}@[\ell]^{\Pi}_{\pi}(\sigma)}
\newcommand{\oprtree}[4]{#3\;\text{\faSitemap}@[\textsf{#1_{#2}}]^{\Pi}_{\pi}(#4)}

\newcommand{\osubtree}[4]{#2\;\text{\faSitemap}@[\textsf{node}_{#1}]^{\Pi}_{\pi}(#3_#4)}
\newcommand{\otree}[3]{#2\;\text{\faSitemap}@[\textsf{node}_{#1}]^{\Pi}_{\pi}(#3)}
\newcommand{\oltree}[3]{#2\;\text{\faSitemap}@[#1]^{\Pi}_{\pi}(#3)}

@@

\paragraph{Virtualization of Disk-Blocks in Filesystems}

\paragraph{Consistency of Disk Layout} \todo{}
\paragraph{Memory Reclamation in Client Code}\todo{iso:speak about Amal's stuff}
\paragraph{Consistency of Reclaimed Memory}\todo{iso: speak about the operator for which the hybrid logic stuff is usefull }

@@

\subsection{Virtualization}
One of the most common forms of abstraction provided by systems software is \emph{virtualization},
which abstracts the relationship between conceptual and physical computing resources. 
Operating System (OS) kernels virtualize memory locations and quantity (via virtual memory and paging~\cite{Denning1970VM}).
Distributed language runtimes may virtualize addresses, even when processes may migrate across machines~\cite{jul1988fine}.
Filesystems virtualize locations on disk~\cite{rodeh2013btrfs,hitz1994file,Rosenblum1992LFS,bonwick2003zettabyte}.
Programs built on top of the corresponding systems software layer work logically at the level of these virtualized resources,
and it makes sense to specify the systems software directly in terms of those abstractions.
\paragraph{Access by Translation}
Accessing virtualized resources via translation is a common way to virtualize notions of location (e.g.,
virtual memory addresses, inodes or object IDs instead of disk addresses~\cite{bonwick2003zettabyte,hitz1994file,Rosenblum1992LFS,rodeh2013btrfs}).
%Accessing the hardware resources in the presence of sharing means having more resource references (references to the virtualized resources) to access the \emph{mapped} resources that are physical and potentially fewer. 
B-trees, page tables, and related structures both behave like maps, when the corresponding physical resources exist as such
just in a different location.
Control over the lookup process (e.g, handling the case of a missing translation entry) allows for additional flexibility,
such as filling holes in sparse files, or demand paging (both from disk or lazily populating anonymous initially-zeroed mappings).
Although the realization of these maps may differ from a system to system based on the context (and sometimes hardware details),
they are semantically --- logically --- partial maps, worth treating as such in verification. 
\paragraph{Modality for Recursive Modal Contexts on Trees}
\label{sec:overviewnode}

To reason about states of trees and subtrees in particular, we work with a modality for stating an assertion that is true \emph{of all nodes within a tree}, identifying the tree by a root note. We refer to the resources that provide the context of these modal assertions as the \emph{modal context}; for a recursively defined tree like those in \textsf{Modfs}, these contexts are recursively defined in terms of the tree structure.

%We define the modal context as the resources (e.g. a btree node tree, free page id list) which are accessed under some relation defined within the modal definition (e.g. being children of a cursor node). The choice of modal context can be structurally dependent on the semantics, in our case \textsc{CoW} semantics.

Write amplification traverses the overall tree depth-first, transforming the tree from the bottom up, one full subtree at a time. Each time a node is split, it results in disconnecting some nodes from the next snapshot, sharing other nodes, and linking some newly-allocated nodes into the new snapshot, as in Figure \ref{fig:cow_index_tree_update}. Nodes are split (or deemed not to require splitting) one at a time, and considered only after all child nodes have been amplified. Concepually this means the amplification code's proof must track the fact that at any point in time, there are regions of the tree where each node has already been amplified (e.g., $\phi_{\textsf{amplified}}$ definitely holds for that node), and regions where nodes may or may not require amplification, and this will not be known until those nodes are visited (i.e., some basic validity $\phi_{\textsf{valid}}$ is known to hold, but not the stronger $\phi_{\textsf{amplified}}$).

%\setlength\intextsep{0pt}
\begin{wrapfigure}{r}{0.25\textwidth}
  \centering  \includegraphics[width=0.25\textwidth]{entail.png}
  \caption{Understanding the level of $\varphi$ satisfaction in the \textsc{CoW} tree of modal nodes}
   \label{fig:nodeentail} 
\end{wrapfigure}
We use a modality similar to $\text{\faSitemap}@[\ell]\phi$ to indicate that $\phi$ is true of the tree node rooted at $\ell$, as well as at \emph{all reachable children} of that node.
This means that establishing a particular $\phi$ throughout a tree must be done in a bottom-up manner (see Figure \ref{fig:nodeentail}).
Capturing this modally, rather than with more typical ad-hoc hand-crafted invariants, offers advantages in a number of proof steps, in addition to concisely capturing intuitive reasoning about subtrees with all nodes satisfying a particular property. One basic example is that while the actual disk sync code requires the full filesystem tree to be in an amplified state (hence write amplification), most of the filesystem works with more general trees. Since $\phi_{\textsf{amplified}}\vdash\phi_{\textsf{valid}}$, using our \faSitemap modality we can conclude $\text{\faSitemap}@[\ell]\phi_{\textsf{amplified}}\vdash\text{\faSitemap}@[\ell]\phi_{\textsf{valid}}$ by an analogue of Axiom K.
There are also other places in the code where a subtree satisfying a stronger assertion at all nodes must be linked into a larger tree satisfying only a weaker assertion overall: predicate representing the valid tree as the tree of nodes which are backed by resources allocated pages, i.e. $\phi_{\textsf{valid}}$, is weaker than the predicate representing the tree of nodes being currently amplified in the current snapshot. Again, the same reasoning principle makes this straightforward. As alluded to earlier, our $\phi_{\textsf{amplified}}$ and company are in fact \emph{predicates} parameterized by a choice of node, though by lifing \textsf{Iris}'s BI structure they can use all of the standard separation logic connectives. The rest of this subsection develops this modality in full detail.
%\colinsaysinl{I cut some text here about the three-way split (commented out) that I wasn't sure how to use, but maybe I deleted something important?}
%To put it concretely, as shown in Figure \ref{fig:cow_index_tree.JPG}, \textsc{CoW} semantics determine what we can claim on the structure of the filesystem indexing tree:
%\begin{itemize}
%\item The bottom-up direction splits the tree into the two half at the cursor node: 1. the bottom part accessible from the cursor node and per-node-amplification is reflected in case needed 2. the upper part on which amplifcation has not reflected yet. 
%\item Horizontally, amplifcation path splits the bottom part of the tree into three different contexts: untouched tree nodes, nodes on the amplification path, and the ones generated as part of node splitting. 
%\end{itemize}
%Based on these two observations, in fact, bottom-up amplification determines the choice of modal context by giving the boundries of \textit{any claimable knowledge} (e.g. $\varphi_{\textsf{amplified}}$, $\varphi_{\textsf{untouched}}$): a subtree bounded by the cursor node.
\subsection{Speculation in Consistency}
%\subsection{Sharing} Sharing is one of the fundamental reasons for kernels to {virtualize} resources,
%for creating the illusion of exclusive access to virtualized versions of shared resources, including CPU time (via preemption).
%An implementation of sharing relies on two concepts: \emph{indirection} and \emph{coordination}. Indirection is a method that allows the mapping of resources of one type into another. By doing so, it allows treating the limited resources (e.g. hardware resources such as physical memory) as if there are more than they actually are (e.g. via virtual memory addresses and paging). Consequently, virtualization for the sake of ensuring availability requires \emph{sharing} of the resources that exist in reality.
%%\paragraph{Isolation Hand-in-Hand with Sharing}
%%\Ismail{put a few words on abstractions such as address space, stack region, snapshots ,etc.}
%%\Colin{Not sure what you want here?}
